State of the Practice for Medical Imaging Software Based on Open Source Repositories

### 2024-11-06 StateOfPractice/Papers/SOP_MI_CritReviewsBioMedEng/SOP_MI_OSR.tex
edit start of abstract and remove all tabs.
start editing

We present the state of the practice for Medical Imaging (MI) software based on
data available in open source repositories. We selected 29 projects from 48
candidates and assessed 9 software qualities (installability, correctness/
verifiability, reliability, robustness, usability, maintainability, reusability,
understandability, and visibility/transparency) by answering 108 questions for
each. Using the Analytic Hierarchy Process (AHP) on the quantitative data, we
ranked the MI software.  The top five are \textit{3D Slicer}, \textit{ImageJ},
\textit{Fiji}, \textit{OHIF Viewer}, and \textit{ParaView}.  This is consistent
with the community's view, with four of these also appearing in the top five
using GitHub metrics (stars-per-year).  The quality and quantity of
documentation present in a project correlates quite well with its popularity.
Generally, MI software is in a healthy state as shown by the following: in the
repositories we observed 88\% of the documentation artifacts recommended by
research software development guidelines, 100\% of MI projects use version
control tools, and developers appear to use the common quasi-agile research
software development process. However, the current state of the practice
deviates from the existing guidelines because of the rarity of some recommended
artifacts (like a test plan, requirements specification, code of conduct, and
code style guidelines), low usage of continuous integration (17\% of the
projects), low use of unit testing (about 50\% of projects), and room for
improvement with documentation. From interviewing the developers, we identified
6 concerns: lack of development time, lack of funding, technology hurdles,
correctness, usability, maintainability, and reproducibility. Recommendations
for improving the state of the practice include the following: increase
documentation, increase testing by enriching datasets, increase continuous
integration usage, move to web applications, employ linters, use peer reviews,
and design for change.

@@

As well as state of the practice (SOP) for MI software, we would like to
understand the impact of the often cited gap (chasm!) between recommended
software engineering practices and that used for most research 
software~\cite{Storer2017}. Although scientists spend
a substantial proportion of their working times on software development
\cite{Hannay2009, Prabhu2011}, few are formally trained~\cite{Hannay2009}.

Our investigation is based on the following four research questions:
\begin{enumerate}
\item[RQ\refstepcounter{rqnum}\therqnum \label{RQ_WhatProjects}:] What MI
{open source} software projects exist? (Section~\ref{SecWhatProjects})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_HighestQuality}:] Based on
quantitative measurements of each project's development practices, which
projects follow best practices? (Section~\ref{SecMeasurementResults})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareHQ2Popular}:] How
similar are the top projects identified in \rqref{RQ_HighestQuality}
to the most popular projects as viewed by the community?
(Section~\ref{Sec_VsCommunityRanking})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareArtifacts}:] How
do MI projects compare to general research software with respect to the
artifacts (documents, scripts and code) present in their repositories?
(Section~\ref{Sec_CompareArtifacts})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareToolsProjMngmnt}:] How
do MI projects compare to research software in general with respect to the use
of tools (Section~\ref{Sec_CompareTools}) for development
(\rqref{RQ_CompareToolsProjMngmnt}.a); and, project management
(\rqref{RQ_CompareToolsProjMngmnt}.b)?
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareMethodologies}:]
How do MI projects compare to research software in general with respect to
principles, processes, and methodologies used?
(Section~\ref{Sec_CompareMethodologies})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_PainPoints}:] What are
the pain points for developers working on MI software projects?
(Section~\ref{painpoints})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_ComparePainPoints}:] How
do the pain points of developers from MI compare to the pain points
for research software in general? (Section~\ref{painpoints})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Concerns}:] For MI
developers what specific best practices are taken to address the pain points
and software quality concerns? (Section~\ref{painpoints})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Recommend}:]
What research software development practice could potentially address the
pain point concerns identified in \rqref{RQ_PainPoints}?
(Section~\ref{ch_recommendations})
\end{enumerate}

@@

\item [Installability] We found installation instructions for 16 projects, but
two did not need them (\textit{BioImage Suite Web} and \textit{Slice:Drop}) as
they are web applications. 10 of the projects required extra dependencies: Five
depend on a specific browser; \textit{dwv}, \textit{OHIF Viewer}, and
\textit{GATE} needs extra libraries to build; \textit{ImageJ} and \textit{Fiji}
need an unzip tool; \textit{MatrixUser} needs Matlab; \textit{DICOM Viewer}
needs a Nextcloud platform. \textit{GATE}, \textit{dwv}, and \textit{DICOM
Viewer} showed severe installation problems. We were not able to install them,
even after a reasonable amount of time (2 hours).

@@

\item [Surface Robustness] According to their documentation, all 29 software
packages should support the DICOM standard. To test robustness, we prepared two
types of image files: correct and incorrect formats (with the incorrect format
created by relabelling a text file to have the ``.dcm'' extension).  All
software packages loaded the correct format image, except for \textit{GATE},
which failed for unknown reasons.  For the broken format, \textit{MatrixUser},
\textit{dwv}, and \textit{Slice:Drop} ignored the incorrect format, did not show
any error message and displayed a blank image.  \textit{MRIcroGL} behaved
similarly except that it showed a meaningless image.  \textit{Drishti}
successfully detected the broken format of the file, but the software crashed as
a result.

@@

To answer \rqref{RQ_CompareHQ2Popular} we use GitHub stars, number of forks and
number of people watching the projects are proxies for community ranking.
Table~\ref{tab_ranking_vs_GitHub} show the statistics for data collected in July
2021.  Recall that 24 projects use GitHub (Section~\ref{SecMeasurementResults}).

@@

Another likely reason for discrepancies is that we weighted all qualities
equally. This is not likely how users implicitly rank the different qualities.
This would require a broader user study to properly assess. Furthermore our
measures of popularity are only \emph{proxies} which are biased towards
past rather than current preferences~\cite{Szulik2017}, as these are
monotonically increasing quantities. Finally there are often more factors than
just quality that influence the popularity of ``consumer'' products.

@@

Although both rankings are imperfect measures, they nevertheless suggest a
correlation between best practices and popularity. We don't know if this is
causal, in either direction (i.e. if best practices enable popularity or if
popularity increases the need for using more software development best
practices).

>>>>

We present the state of the practice for the development of Medical Imaging
(MI) software based on data available in open source repositories. We selected
29 projects from 48 candidates and assessed 9 software qualities
(installability, correctness/ verifiability, reliability, robustness,
usability, maintainability, reusability, understandability, and
visibility/transparency) by answering 108 questions for each. Using the
Analytic Hierarchy Process (AHP) on the quantitative data, we ranked the MI
software.  The top five are \textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji},
\textit{OHIF Viewer}, and \textit{ParaView}.  This is consistent with the
community's view, with four of these also appearing in the top five using
GitHub metrics (stars-per-year).  The quality and quantity of documentation
present in a project correlates quite well with its popularity.  Generally, MI
software is in a healthy state: in the repositories
we observed 88\% of the documentation artifacts recommended by research
software development guidelines, 100\% of MI projects use version control
tools, and developers appear to use the common quasi-agile research software
development process. However, the current state of the practice deviates from
existing guidelines as some recommended artifacts are rarely present
(like a test plan, requirements specification, code of conduct, and code style
guidelines), low usage of continuous integration (17\% of the projects), low
use of unit testing (about 50\% of projects), and room for improvement with
documentation. From developper interviews , we identified 7 concerns: lack
of development time, lack of funding, technology hurdles, correctness,
usability, maintainability, and reproducibility. We recommend:
increase effort on documentation,
increase testing by enriching datasets, increase continuous integration usage,
move to web applications, employ linters, use peer reviews, and design for
change.

@@

As well as state of the practice (SOP) for MI software, we would like to
understand the impact of the often cited gap between recommended
software engineering practices and that used for most research 
software~\cite{Storer2017}. Although scientists spend
a substantial proportion of their working times on software development
\cite{Hannay2009, Prabhu2011}, few are formally trained~\cite{Hannay2009}.

Our investigation is based on the following research questions:
\begin{enumerate}
\item[RQ\refstepcounter{rqnum}\therqnum \label{RQ_WhatProjects}:] What MI
{open source} software projects exist? (Section~\ref{SecWhatProjects})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_HighestQuality}:] Based on
quantitative measurements of each project's development practices, which
projects follow best practices? (Section~\ref{SecMeasurementResults})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareHQ2Popular}:] How
similar are the top projects identified in \rqref{RQ_HighestQuality}
to the most popular projects as viewed by the community?
(Section~\ref{Sec_VsCommunityRanking})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareArtifacts}:] How
do MI projects compare to general research software with respect to the
artifacts (documents, scripts and code) present in their repositories?
(Section~\ref{Sec_CompareArtifacts})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareToolsProjMngmnt}:] How
do MI projects compare to research software in general with respect to the use
of tools (Section~\ref{Sec_CompareTools}) for development
(\rqref{RQ_CompareToolsProjMngmnt}.a); and, project management
(\rqref{RQ_CompareToolsProjMngmnt}.b)?
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareMethodologies}:]
How do MI projects compare to research software in general with respect to
principles, processes, and methodologies used?
(Section~\ref{Sec_CompareMethodologies})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_PainPoints}:] What are
the pain points for developers working on MI software projects?
(Section~\ref{painpoints})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_ComparePainPoints}:] How
do the pain points of developers from MI compare to the pain points
for research software in general? (Section~\ref{painpoints})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Concerns}:] For MI
developers what specific best practices are taken to address the pain points
and software quality concerns? (Section~\ref{painpoints})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Recommend}:]
What research software development practice could potentially address the
pain point concerns identified in \rqref{RQ_PainPoints}?
(Section~\ref{ch_recommendations})
\end{enumerate}

@@

\item [Installability] We found installation instructions for 16 projects, but
two did not need them (\textit{BioImage Suite Web} and \textit{Slice:Drop}) as
they are web applications. 10 of the projects required extra dependencies: Five
depend on a specific browser; \textit{dwv}, \textit{OHIF Viewer}, and
\textit{GATE} needs extra libraries to build; \textit{ImageJ} and \textit{Fiji}
need an unzip tool; \textit{MatrixUser} needs Matlab; \textit{DICOM Viewer}
needs a Nextcloud platform. We were not able to install \textit{GATE},
\textit{dwv}, \textit{DICOM Viewer} even after trying for 2 hours (each).

@@

\item [Surface Robustness] According to their documentation, all 29 software
packages should support the DICOM standard. To test robustness, we prepared two
types of image files: correct and incorrect formats (with the incorrect format
created by relabelling a text file to have the ``.dcm'' extension).  All
software packages loaded the correct format image, except for \textit{GATE},
which failed for unknown reasons.  For the broken format, \textit{MatrixUser},
\textit{dwv}, and \textit{Slice:Drop} ignored the incorrect format, did not show
any error message and displayed a blank image.  \textit{MRIcroGL} behaved
similarly except that it showed a meaningless image.  \textit{Drishti}
successfully detected the broken format, but the software then crashed.

@@

To answer \rqref{RQ_CompareHQ2Popular} we use GitHub stars, number of forks and
number of people watching the projects as proxies for community ranking.
Table~\ref{tab_ranking_vs_GitHub} show the statistics for data collected in July
2021.

@@

We weighted all qualities equally, which is not the likely (implicit) weight
that users implicitly use. To properly assess this would require a broad user
study.  Furthermore our measures of popularity are only \emph{proxies} which
are biased towards past rather than current preferences~\cite{Szulik2017}, as
these are monotonically increasing quantities. Finally there are often more
factors than just quality that influence the popularity of ``consumer''
products.

@@

Although both rankings are imperfect measures, they nevertheless suggest a
correlation between best practices and popularity. We don't know if this is
causal, in either direction (i.e. if best practices enable popularity or if
popularity increases the need for best practices).
